### 2021-07-05 ch/2-related-work.tex
More corrections

The authors define the criteria in terms of statistical independence:
\acf{EOdds} refers to the case where \(\hat{y}\) and \(s\) are independent conditional on \(y\).
This is equivalent to the true positive rate and the false positive rate being the same for all groups.
\ac{EOpp} is, as mentioned above,
the case where just the true positive rate is the same for all groups.
This means that \(\hat{y}\) and \(s\) are independent conditional on \(y=1\).
One main advantage that these definitions have over \ac{DP} is that a perfect predictor satisfies them on any evaluation set.
% Demographic Parity is at odds with achieving perfect accuracy.
% Furthermore, for Demographic Parity a classifier might be forced to give a positive prediction to a lot of individuals
% where it is really not appropriate.

In order to construct a fair classifier out of a \emph{binary predictor}
(giving only binary predictions) via post-processing,
the output is randomised in such a way as to remove the bias.
If the overall probability for a positive predictions of the unfair classifier is given by \(P(\hat{y}=1|s)\),
then the randomised probability for the fair positive predictions \(\tilde{y} = 1\) is given by:
\begin{align}
  \label{eq:hardt}
  P(\tilde{y}=1| x, s) = \sum\limits_{y' \in \{0, 1\}} P(\tilde{y} = 1| \hat{y}=y', s) 
  P(\hat{y}=y'| x, s)
\end{align}
There are 4 free parameters \(P(\tilde{y} = 1| \hat{y}=y', s=s')\) with \(y' \in \{0, 1\}\) and \(s' \in \{0, 1\}\).
As an example, if the unfair predictor predicts \(\hat{y} =0\) for an input with \(s=0\),
then the fair predictor predicts \(\tilde{y} =1\) with probability \(P(\tilde{y} = 1| \hat{y}=0, s=0)\)
which might be non-zero.
In addition to the fairness condition, we also want to enforce accuracy.
To that end, we introduce a loss function \(\ell (\tilde{y}, y)\)
that quantifies the cost of predicting the wrong class label.
The final optimisation problem for the 4 free parameters is then to minimise \(\ell\)
under the constraint of \ac{EOpp} or \ac{EOdds}
and the constraint that the parameters must be valid probabilities.

For a predictor that outputs a \emph{score function},
the post-processing step consists of choosing differing thresholds for \(s=0\) and \(s=1\),
such that the predictions become fair.
If \(f\) is the score, then for a given threshold \(t\) we predict \(\hat{y} = 1\) if \(f > t\).
The two thresholds (one for each group) are found by minimising \(\ell\) with the fairness constraints, as before.
To satisfy the constraint, it might be necessary to randomise the result.
This is done by using two constraints per group:
if \(f\) is above both, the result is \(\hat{y} = 1\), if it is below both, \(\hat{y} =0\)
and if \(f\) is between the thresholds, then the result is chosen at random.
This method -- as well as the one before --
requires the sensitive attribute for all predictions as it has to choose the right threshold.

@@

\subsection{Fair classifiers continued}
Following the initial publications,
several works refined the ideas that had been presented there.
A very influential work by \citet{zafar2017fairnessconstraints}
presents a more efficient way to train a fair classifier.
The authors explicitly discuss the tension
between the two goals of making fair predictions and not making use of the sensitive attributes.
Using sensitive attributes to make predictions
is referred to as \emph{Disparate Treatment} by the authors.
Algorithms like the one by \citet{calders2009building} use the sensitive
attribute during prediction to achieve a very high degree of fairness, but this can have a
lot of problems from a legal perspective and can also easily go wrong.
(Making use of \(s\) comes close to \emph{direct discrimination} discussed above.)
Thus, \citet{zafar2017fairnessconstraints} set themselves the goal of achieving fairness while making use of the sensitive attributes
as little as possible, i.e.~avoiding Disparate Treatment.

In order to train a fair classifier, \citet{zafar2017fairnessconstraints} use a proxy of the definition of
\ac{DP} that is easier to optimise during training, namely the covariance
between the sensitive attribute and the predicted score \(f\):
\begin{align}
  \label{eq:zafar-constraint}
  Cov(s, f) = \mathbb{E}[(s - \bar{s})f] - \mathbb{E}[(s - \bar{s})]\bar{f} \approx 
  \frac{1}{N} \sum\limits_{i=1}^{N} (s_i - \bar{s}) \cdot f_i
\end{align}
The authors refer to this as the covariance criterion.
The classifier can then be trained in two ways.
The first way is to minimise the loss that enforces accuracy
with the constraint that the covariance criterion is below a certain threshold.
The other way is to minimise the covariance criterion
with the constraint that the accuracy is above a certain threshold.
The paper includes experiments with these techniques with \ac{SVM} and Logistic Regression classifiers
on the Adult/Census Income dataset and the Bank marketing dataset~\citep{Dua:2017}.
In the Bank marketing dataset, the task is to predict whether a client will respond to marketing for a term deposit,
based on 20 attributes of the person.
The sensitive attribute is \emph{age}.
The proposed algorithm performs similarly to the best previous algorithms (similar in fairness and accuracy),
notably without making use of the sensitive attribute for predictions.
% This is an impressive achievement.

In a follow up work, \citet{zafar2017fairnesstreatment} adapt their work to avoid ``disparate mistreatment''.
Avoiding disparate mistreatment is equivalent to enforcing \acf{EOdds}
as defined in the contemporary work by \citet{hardt2016equality}.
As in the previous work, they define a tractable proxy for this measure.
To this end, they first define a function \(g\) that measures
how well the class label \(y \in \{0, 1\}\) and the predicted score for a data point \(0 \leq f(x) \leq 1\) agree:
\begin{align}
  \label{eq:zafar-constraint-2}
  g(y, x) = \min \left(0, \left(y - \tfrac{1}{2}\right) \cdot \left(f(x) - \tfrac{1}{2}\right)\right)
\end{align}
Subtracting \(\tfrac{1}{2}\) is necessary to centre the values around \(0\).
The proxy for \ac{EOdds} is then the covariance between the sensitive attribute and \(g\).
\(g\) can also be modified to only take into account cases with \(y=1\) which corresponds to Equality of Opportunity.
The paper contains experiments on the ProPublica/COMPAS dataset~\citep{angwin2016machine}.
The task for this dataset is to predict
whether a criminal offender committed another misconduct or felony within two years,
based on personal information that includes age, gender, race and past criminal history.
% The authors compare the proposed method to an implementation of the algorithm from \citet{hardt2016equality}.
% In the Equality of Opportunity task, the approach by \citet{hardt2016equality} performs better
% (lower discrimination, similar accuracy).
% The better performance is likely at least partially explained
% by \citet{hardt2016equality} using the sensitive attribute for predictions.
% The evaluation metric is
% the difference in FPR (false positive rate) and the difference in FNR (false negative rate) between the two groups.
% Hardt et al.~achieve an FPR difference of 0.02 at 66\% accuracy while Zafar et al.~achieve an FPR difference of
% 0.06 at the same 66\% accuracy.
% For Equalised Odds,
% the proposed approach achieves higher accuracy but also higher discrimination than in \citet{hardt2016equality}.
% One limiting factor that the authors identify is the size of the dataset:
% the degree of discrimination cannot be accurately estimated by the algorithm if the dataset is small.

\citet{woodworth2017learning} also consider \ac{EOdds} as a fairness definition
but give strong criticism for the post-processing approach from \citet{hardt2016equality}.
They present examples where finding the optimal (but discriminatory) predictor in a particular hypothesis class
and then correcting it post-hoc performs poorly.
% The proofs are based on the PAC learning framework.
Certain training distributions allow the training of good fair classifiers from scratch
but when doing post-processing on a Bayes-optimal predictor for this distribution,
the result can be a very bad predictor that gives essentially random predictions.
The authors explicitly construct one such distribution:
it corresponds to the case where the sensitive attribute \(s\) is more predictive of the class label \(y\)
than the features \(x\), which is arguably an unusual case.

@@

This topic has since received more contributions than can be listed here.
The following is a very short sample.
In the direction of \citet{zafar2017fairnessconstraints}, there is for example \citet{quadrianto2017recycling,ustun2019fairness,lohaus2020too}.
In the direction of \citet{kamiran2012data}, there is for example \citet{AgaBeyDudLanetal18,roh2021fairbatch}.
In the direction of \citet{hardt2016equality}, there is for example \citet{kim2019multiaccuracy}.

@@

This is the premise of \citet{edwards2016censoring}.
Four neural networks are trained in an adversarial setting:
the encoder \(f(x)\) which encodes the unfair representation \(x\) into a fair representation \(z\),
the classifier \(g(z)\) which tries to predict \(y\) from \(z\),
the decoder \(k(z)\) which tries to reconstruct \(x\) from \(z\),
and finally the adversary \(h(z)\) which tries to predict \(s\) from \(z\).
The classifier \(g\) ensures that \(z\) contains enough information to predict \(y\).
The decoder \(k\) ensures that the fair representation contains enough information from \(x\)
and can be used by other classifiers as well.
The sign of the adversary's loss is inverted for the gradient of the encoder \(f\),
so that encoder tries to make the adversary's predictions worse.
Encoder and adversary should converge to a point
where the adversary can only predict the correct \(s\) from \(z\) at chance level.
Assuming that the adversary is powerful enough to detect any trace of \(s\) in \(z\),
this means that no information about the sensitive attribute \(s\) remains in the fair representation \(z\).
Note that there is a tension here between requiring \(x\) to be reconstructible from \(z\)
and requiring \(z\) to contain no information about \(s\), since \(x\) \emph{does} contain information about \(s\).

@@

A contemporary paper by \citet{kusner2017counterfactual} approaches the problem differently.
In this paper, a new fairness criterion which they call ``counterfactual fairness''
is defined making use of causal reasoning.
For the definition, they assume access to a causal model
that can be used to compute \emph{counterfactuals} (for example a structural equation model \citep{kaplan2008structural}).
This is in contrast to \citet{kilbertus2017avoiding} who relied on just \emph{interventions}.
(According to Pearl's causal hierarchy \citep{pearl2019seven},
counterfactuals require a deeper level of causal information than interventions.)
The causal model depends on unobserved background variables \(U\),
non-sensitive features \(x\) and a sensitive attribute \(s\).
A predictor \(\hat{y}\) is then \emph{counterfactually fair}
if the following equality of counterfactual probabilities holds:
\begin{align}
  \label{eq:counterfactual-fairness}
  P(\hat{y}_{s=i}(U) = 1|x, s=i)=P(\hat{y}_{s=j}(U) = 1|x, s=i)
\end{align}
where \(i, j \in \{0, 1\}\) and \(i \neq j\).

The first counterfactual probability is actually not a real counterfactual:
it is just the probability of observing \(\hat{y} = 1\)
given the background variables \(U\), the features \(x\) and \(s=i\).
The second counterfactual probability is the probability of observing \(\hat{y} = 1\)
given \(U\) and \(x\) and given that we had \(s=j\) instead of the actual value \(s=i\).
In other words, the criterion demands this:
the prediction would have been the same in the counterfactual world
where sensitive attribute is \(j\) instead of \(i\).
In the counterfactual world, everything that is not causally dependent on \(s\) is held constant.

@@

The work by \citet{chiappa2019path} can be seen as a refinement of \citet{kilbertus2017avoiding}.
Instead of disallowing influence of \emph{nodes}, this works disallows certain \emph{paths} to influence the outcome.
The path pointing directly from the sensitive attribute to the outcome is not allowed,
but paths via certain variables are allowed.
An example for such a permissible variable comes from the Berkeley admissions dataset~\citep{bickel1975sex}:
overall, the data shows that women were admitted at lower rates,
but this turned out to be mediated by department choice.
Women were applying to more competitive departments and thus had lower acceptance rates.
The choice of department \(C\) is here an admissible variable:
the outcome \(Y\) may depend on it.
So, a path from \(S\) (gender) to \(Y\) via \(C\) is permissible, and our fairness definition should reflect this.
Counterfactual fairness \citep{kusner2017counterfactual} would in this case arguably give the wrong answer.
In certain simple cases, path-specific fairness is equivalent to the criterion from \citet{kilbertus2017avoiding}.
The problem remains that the causal model needs to be constructed
and permissible variables need to be identified manually.

@@

\section{Ground-truth-centric view of bias}%
\label{sec:groundtruth-centric-view-of-bias}
The fairness-accuracy trade-off which results from fairness constraints like \ac{DP} and \ac{EOpp}
has often been seen as a thorny issue.
\citet{wick2019unlocking} were one of the first to argue for a change of perspective to solve this.
For example, the labels of the ProPublica/COMPAS~\citep{angwin2016machine} dataset had been treated as unbiased ground truth in previous work,
which the authors call into question.
The simple change they do is evaluating fairness-enforcing models on actual \emph{unbiased} test data.
In practice, they simulate datasets such that unbiased labels can be used for evaluation,
which is necessary because they do not have access to (unobserved) ground truth labels.
The paper considers label bias and ``selection bias''
(``selecting a subsample of the data in such a way that happens to introduce unexpected correlations,
say, between a protected attribute and the target label''),
which we previously referred to as sample bias.
In their experiments they find,
that enforcing fairness can in some situations improve the accuracy on the (unbiased) test set.

\citet{kim2019learning} also consider the problem of learning from biased data (and evaluating on unbiased data).
While they do not formulate the problem as a fairness problem,
it is equivalent to enforcing fairness in the presence of severe sampling bias.
Their main motivating example is the Coloured MNIST dataset;
a dataset derived from the MNIST dataset~\citep{lecun1994mnist}.
In this dataset, digits are randomly coloured in the test set,
but in the training set there is a one-to-one correspondence between digit class and colour.
A na\"ive classifier will learn to predict colour instead of digit class.
The proposed method is very similar to \citet{ganin2016domain} and \citet{edwards2016censoring},
if we think of the colours as forming different domains
(\ie, the \emph{red} domain contains all red digits, etc.).
The goal is then to learn a domain-independent representation.
The main difference to previous work is an additional entropy loss term in the adversarial loss.

\citet{arjovsky2019invariant} tackle a similar problem, but take a very different approach,
which they term ``Invariant Risk Minimisation'', in contrast to the default method \acf{ERM}.
As with \citet{kim2019learning},
the goal is to ignore spurious correlations that only appear in the very imperfect training set,
but not in the test set;
and they also perform experiments on Coloured MNIST (see above).
They formalise the problem as one of different environments $e\in E$,
where each environment is a different, biased view of the same underlying data distribution.
The idea is to train a predictor that is simultaneously optimal for all the environments,
with the expectation that this generalises to the test set.
This idea is formulated as a (intractable) bi-level, constrained optimisation problem,
which they approximate with a tractable regularised optimisation.

As the authors point out,
the goal of machine learning should be to identify natural, deep, robust structures in reality,
instead of relying on superficial correlations.
(This dichotomy is sometimes framed as \emph{correlation} vs \emph{causality},
where ``causality'' takes on a very broad meaning that encompasses any kind of fundamental structure in reality.)
Methods for invariance learning can be framed as trying to learn a more fundamental structure
than the dataset at first seems to show.
By aiming to be invariant to specific environments/subgroups/domains (denoted by the variable \(s\)),
they can be seen as making the claim that \(s\) does not represent a natural structure,
but is merely an artefact of the data generation process.
Furthermore, the ground-truth-centric view of dataset bias can be understood as
aiming to identify the fundamental structures and aiming to be invariant to everything else.
This also fits with \citet{kim2019learning}'s work on Coloured MNIST:
colour is not regarded as the fundamental structure in the data.

@@

A conceptually very influential work was \citet{friedler2016possibility}.
While this work does not explicitly present a ground-truth-centric approach to bias,
their concepts of the construct space and the observed space are close
to the idea of the (biased) training distribution and the true underlying distribution.

@@

\citet{kallus2018residual} is another more conceptual work.
They present an intuitive model for how sampling bias can enter a training set
and how it can be very hard to correct for it.
The paper refers to the process as \emph{systematic censoring},
but this is not meant to necessarily imply that this censoring was a conscious decision by some authority;
it can also be an unintended side effect of an enacted policy.
Systematic censoring can arise any time a screening process prevents observing the outcome for the screened-out samples.
For example, if, historically, a certain demographic group was screened out from receiving loans,
then it was not possible to observe default rates for this group;
and so any historical data on defaulting is useless for that demographic group.
Even if a bank wants to change their policy,
it is hard to correct for the bias in the training set because there is not good data to learn from.
This is a prime example of a ground-truth-centric bias problem.

Finally, as mentioned in the beginning,
\citet{blum2020recovering} provide a formalisation of a setting with label bias and sampling bias
and an unbiased ground truth test set.
They explicitly state not being affected by a fairness-accuracy trade-off,
because the true distribution in the considered problem satisfies \acf{DP}.
Their formalised dataset generation process has multiple steps:
First, a small amount of random noise is applied to the (true) labels;
this models general inaccuracies in the labelling process and does not introduce bias yet.
Second, sampling bias is added, by dropping samples based on what the sample's \(s\) and \(y\) values are.
Concretely, only samples with \(s=0\) are dropped.
Finally, labelling bias is added, by flipping labels for one of the demographic groups (\eg, \(s=0\)),
in one specific direction (from \(y=1\) to \(y=0\)).
The authors show that in this specific situation,
an unbiased classifier (as determined by evaluation on the unbiased test set) can be recovered
by enforcing its predictions to be compatible with \acf{EOpp} on the training set.
Notably, the fairness constraint enforced during training (\ac{EOpp})
is not the fairness constraint satisfied by the predictions on the unbiased test set (\ac{DP}).
One reason, why enforcing \ac{EOpp} works so well here is
that the label bias only affects the \emph{underrepresented} demographic group,
and that the label bias is the direction (\(y=1\) to \(y=0\)) that \ac{EOpp} is most sensitive to.
Thus, even in the ground-truth-centric view of bias,
fairness constraints can be very useful.

>>>>

In this ``oblivious'' setting, the authors define their fairness criteria in terms of statistical independence:
\acf{EOdds} refers to the case where \(\hat{y}\) and \(s\) are independent conditional on \(y\).
This is equivalent to the true positive rate and the false positive rate being the same for all groups.
\ac{EOpp} is, as mentioned above,
the case where just the true positive rate is the same for all groups.
This means that \(\hat{y}\) and \(s\) are independent conditional on \(y=1\).
One of the main advantages of these definitions compared to \ac{DP} is that a perfect predictor satisfies them on any evaluation set.
% Demographic Parity is at odds with achieving perfect accuracy.
% Furthermore, for Demographic Parity a classifier might be forced to give a positive prediction to a lot of individuals
% where it is really not appropriate.

In order to construct a fair classifier out of a \emph{binary predictor}
(giving only hard binary predictions $\{0, 1\}$) via post-processing,
the output is randomised in such a way as to remove the bias.
If the overall probability for a positive prediction of the unfair classifier is given by \(P(\hat{y}=1|s)\),
then the randomised probability for the fair positive predictions \(\tilde{y} = 1\) is given by:
\begin{align}
  \label{eq:hardt}
  P(\tilde{y}=1| x, s) = \sum\limits_{y' \in \{0, 1\}} P(\tilde{y} = 1| \hat{y}=y', s) 
  P(\hat{y}=y'| x, s)
\end{align}
There are 4 free parameters \(P(\tilde{y} = 1| \hat{y}=y', s=s')\) with \(y' \in \{0, 1\}\) and \(s' \in \{0, 1\}\).
As an example, if the unfair predictor predicts \(\hat{y} =0\) for an input with \(s=0\),
then the fair predictor predicts \(\tilde{y} =1\) with probability \(P(\tilde{y} = 1| \hat{y}=0, s=0)\)
which might be non-zero.
In addition to the fairness condition, we also want to enforce accuracy.
To that end, the authors introduce a loss function \(\ell (\tilde{y}, y)\)
that quantifies the cost of predicting the wrong class label.
The final optimisation problem for the 4 free parameters is then to minimise \(\ell\)
under the constraint of \ac{EOpp} or \ac{EOdds}
and the constraint that the parameters must be valid probabilities.

For a predictor that outputs a \emph{score function},
the post-processing step consists of choosing differing thresholds for \(s=0\) and \(s=1\),
such that the predictions become fair.
If \(f\) is the score, then for a given threshold \(t\) we predict \(\hat{y} = 1\) if \(f > t\).
The two thresholds (one for each group) are found by minimising \(\ell\) with the fairness constraints, as before.
To satisfy the constraint, it might be necessary to randomise the result.
This is done by using two constraints per group:
if \(f\) is above both, the result is \(\hat{y} = 1\), if it is below both, \(\hat{y} =0\)
and if \(f\) is between the thresholds, then the result is chosen at random.
This method -- as well as the one before --
requires the sensitive attribute for all predictions at test time, as it has to choose the right threshold.

@@

\subsection{Fair classifiers continued}
Following the initial publications,
several works refined the ideas that had been presented there.
A very influential work by \citet{zafar2017fairnessconstraints}
presents a more efficient way to train a fair classifier.
The authors explicitly discuss the tension
between the two goals of making fair predictions and not making use of the sensitive attributes at test time.
Using sensitive attributes to make predictions
is referred to as \emph{Disparate Treatment} by the authors.
Algorithms like the one by \citet{calders2009building} use the sensitive
attribute during prediction to achieve a very high degree of fairness, but this can have a
lot of problems from a legal perspective and can also easily go wrong.
(Making use of \(s\) comes close to \emph{direct discrimination} discussed above.)
Thus, \citet{zafar2017fairnessconstraints} set themselves the goal of achieving fairness while making use of the sensitive attributes
as little as possible, i.e.~avoiding Disparate Treatment.

In order to train a fair classifier, \citet{zafar2017fairnessconstraints} use a proxy of the definition of
\ac{DP} that is easier to optimise during training, namely the covariance
between the sensitive attribute and the predicted score \(f\):
\begin{align}
  \label{eq:zafar-constraint}
  Cov(s, f) = \mathbb{E}[(s - \bar{s})f] - \mathbb{E}[(s - \bar{s})]\bar{f} \approx 
  \frac{1}{N} \sum\limits_{i=1}^{N} (s_i - \bar{s}) \cdot f_i
\end{align}
The authors refer to this as the covariance criterion.
The classifier can then be trained in two ways.
The first way is to minimise the loss that enforces accuracy
with the constraint that the covariance criterion is below a certain threshold.
The other way is to minimise the covariance criterion
with the constraint that the accuracy is above a certain threshold.
The paper includes experiments with these techniques with \ac{SVM} and \acf{LR} classifiers
on the Adult/Census Income dataset and the Bank marketing dataset~\citep{Dua:2017}.
In the Bank marketing dataset, the task is to predict whether a client will respond to marketing for a term deposit,
based on 20 attributes of the person.
The sensitive attribute is \emph{age}.
The proposed algorithm performs similarly to the best previous algorithms (similar in fairness and accuracy),
notably without making use of the sensitive attribute for predictions.
% This is an impressive achievement.

In a follow-up work, \citet{zafar2017fairnesstreatment} adapt their work to avoid ``disparate mistreatment''.
Avoiding disparate mistreatment is the same as enforcing \acf{EOdds},
as defined in the contemporary work by \citet{hardt2016equality}.
As in the previous work, they define a tractable proxy for this measure.
To this end, they first define a function \(g\) that measures
how well the class label \(y \in \{0, 1\}\) and the predicted score for a data point \(0 \leq f(x) \leq 1\) agree:
\begin{align}
  \label{eq:zafar-constraint-2}
  g(y, x) = \min \left(0, \left(y - \tfrac{1}{2}\right) \cdot \left(f(x) - \tfrac{1}{2}\right)\right)
\end{align}
Subtracting \(\tfrac{1}{2}\) is necessary to centre the values around \(0\).
The proxy for \ac{EOdds} is then the covariance between the sensitive attribute and \(g\).
\(g\) can also be modified to only take into account cases with \(y=1\) which corresponds to enforcing \acf{EOpp.
The paper contains experiments on the ProPublica/COMPAS dataset~\citep{angwin2016machine}.
The task for this dataset is to predict
whether a criminal offender committed another misconduct or felony within two years,
based on personal information that includes age, gender, race and past criminal history.
% The authors compare the proposed method to an implementation of the algorithm from \citet{hardt2016equality}.
% In the Equality of Opportunity task, the approach by \citet{hardt2016equality} performs better
% (lower discrimination, similar accuracy).
% The better performance is likely at least partially explained
% by \citet{hardt2016equality} using the sensitive attribute for predictions.
% The evaluation metric is
% the difference in FPR (false positive rate) and the difference in FNR (false negative rate) between the two groups.
% Hardt et al.~achieve an FPR difference of 0.02 at 66\% accuracy while Zafar et al.~achieve an FPR difference of
% 0.06 at the same 66\% accuracy.
% For Equalised Odds,
% the proposed approach achieves higher accuracy but also higher discrimination than in \citet{hardt2016equality}.
% One limiting factor that the authors identify is the size of the dataset:
% the degree of discrimination cannot be accurately estimated by the algorithm if the dataset is small.

\citet{woodworth2017learning} also consider \ac{EOdds} as a fairness definition
but give strong criticism for the post-processing approach from \citet{hardt2016equality}.
They present examples where finding the optimal (but discriminatory) predictor in a particular hypothesis class
and then correcting it post-hoc performs poorly.
% The proofs are based on the PAC learning framework.
Certain training distributions allow the training of good fair classifiers from scratch
but when doing post-processing on a Bayes-optimal predictor for this distribution,
the result can be a very bad predictor that gives essentially random predictions.
The authors explicitly construct one such distribution:
it corresponds to the case where the sensitive attribute \(s\) is more predictive of the class label \(y\)
than the features \(x\),
which becomes an especially severe problem when \(s\) is only predictive during training time, but not at test time.
See the discussion of \emph{Coloured MNIST} below for such an example.

@@

This topic has since received more contributions than can be listed here.
The following is a very short sample.
In the direction of \citet{zafar2017fairnessconstraints}, there is for example \citet{quadrianto2017recycling,ustun2019fairness,lohaus2020too}.
In the direction of \citet{kamiran2012data}, there is for example \citet{AgaBeyDudLanetal18,roh2021fairbatch}.
In the direction of \citet{hardt2016equality}, there is for example \citet{hebert2018multicalibration}.

@@

This is the premise of \citet{edwards2016censoring}.
Four neural networks are trained in an adversarial setting:
the encoder \(f(x)\) which encodes the unfair representation \(x\) into a fair representation \(z\),
the classifier \(g(z)\) which tries to predict \(y\) from \(z\),
the decoder \(k(z)\) which tries to reconstruct \(x\) from \(z\),
and finally the adversary \(h(z)\) which tries to predict \(s\) from \(z\).
The classifier \(g\) ensures that \(z\) contains enough information to predict \(y\).
The decoder \(k\) ensures that the fair representation contains enough information from \(x\)
and can be used by other classifiers as well.
The sign of the adversary's loss is inverted for the gradient of the encoder \(f\),
so that the encoder tries to make the adversary's predictions worse.
Encoder and adversary should converge to a point
where the adversary can only predict the correct \(s\) from \(z\) at chance level.
Assuming that the adversary is powerful enough to detect any trace of \(s\) in \(z\),
this means that no information about the sensitive attribute \(s\) remains in the fair representation \(z\).
Note that there is a tension here between requiring \(x\) to be reconstructible from \(z\)
and requiring \(z\) to contain no information about \(s\), since \(x\) \emph{does} contain information about \(s\).

@@

A contemporary paper by \citet{kusner2017counterfactual} approaches the problem differently.
In this paper, a new fairness criterion which they call ``counterfactual fairness''
is defined making use of causal reasoning.
For the definition, they assume access to a causal model
that can be used to compute \emph{counterfactuals} (for example a structural equation model \citep{kaplan2008structural}).
This is in contrast to \citet{kilbertus2017avoiding} who relied on just \emph{interventions}.
(According to Pearl's causal hierarchy \citep{pearl2019seven},
counterfactuals require a deeper level of causal information than interventions.)
The causal model depends on unobserved background variables \(U\),
non-sensitive features \(x\) and a sensitive attribute \(s\).
A predictor \(\hat{y}\) is then \emph{counterfactually fair}
if the following equality of \emph{counterfactual probabilities} holds:
\begin{align}
  \label{eq:counterfactual-fairness}
  P(\hat{y}_{s=i}(U) = 1|x, s=i)=P(\hat{y}_{s=j}(U) = 1|x, s=i)
\end{align}
where \(i, j \in \{0, 1\}\) and \(i \neq j\).
The first counterfactual probability is actually not a real counterfactual:
it is just the probability of observing \(\hat{y} = 1\)
given the background variables \(U\), the features \(x\) and \(s=i\).
The second counterfactual probability is the probability of observing \(\hat{y} = 1\)
given \(U\) and \(x\) and given that we had \(s=j\) instead of the actual value \(s=i\).
In other words, the criterion demands this:
the prediction would have been the same in the counterfactual world
where the sensitive attribute is \(j\) instead of \(i\).
In the counterfactual world, everything that is not causally dependent on \(s\) is held constant.

@@

\citet{chiappa2019path} have a similar approach,
but their work by  can be seen as a refinement of \citet{kilbertus2017avoiding}.
Instead of disallowing influence of \emph{nodes}, this work disallows certain \emph{paths} to influence the outcome.
The path pointing directly from the sensitive attribute to the outcome is not allowed,
but paths via certain variables are allowed.
An example for such a permissible variable comes from the Berkeley admissions dataset~\citep{bickel1975sex}:
overall, the data shows that women were admitted at lower rates,
but this turned out to be mediated by department choice.
Women were applying to more competitive departments and thus had lower acceptance rates.
The choice of department \(C\) is here an admissible variable:
the outcome \(Y\) may depend on it.
So, a path from \(S\) (gender) to \(Y\) via \(C\) is permissible, and our fairness definition should reflect this.
Counterfactual fairness \citep{kusner2017counterfactual} would in this case arguably give the wrong answer.
In certain simple cases, path-specific fairness is equivalent to the criterion from \citet{kilbertus2017avoiding}.
The problem remains that the causal model needs to be constructed
and permissible variables need to be identified manually.

@@

\section{Ground-truth-centric view of bias}%
\label{sec:groundtruth-centric-view-of-bias}
The fairness-accuracy trade-off which results from fairness constraints like \ac{DP} and \ac{EOpp}
has often been seen as a thorny issue.
\citet{wick2019unlocking} were one of the first to argue for a change of perspective to solve this.
For example, the labels of the ProPublica/COMPAS~\citep{angwin2016machine} dataset had been treated as unbiased ground truth in previous work,
which the authors call into question.
The simple change they do is evaluating fairness-enforcing models on actual \emph{unbiased} test data.
In practice, they use simulated datasets such that unbiased labels can be used for evaluation,
which is in contrast to real data where we do not have access to ground truth labels,
and cannot evaluate on them.
The paper considers label bias and ``selection bias''
(``selecting a subsample of the data in such a way that happens to introduce unexpected correlations,
say, between a protected attribute and the target label''),
which we previously referred to as sample bias.
In their experiments they find,
that enforcing fairness can in some situations improve the accuracy on the (unbiased) test set.

\citet{kim2019learning} also consider the problem of learning from biased data (and evaluating on \emph{un}biased data).
While they do not formulate the problem as a fairness problem,
it is equivalent to enforcing fairness in the presence of severe sampling bias.
Their main motivating example is the Coloured MNIST dataset;
a dataset derived from the MNIST dataset~\citep{lecun1994mnist}.
In this dataset, digits are randomly coloured in the test set,
but in the training set there is a one-to-one correspondence between digit class and colour.
A na\"ive classifier will learn to predict colour instead of digit class.
The proposed method is very similar to \citet{ganin2016domain} and \citet{edwards2016censoring},
if we think of the colours as forming different domains
(\ie, the \emph{red} domain contains all red digits, etc.).
The goal is then to learn a domain-independent representation.
The main difference to previous work is an additional entropy loss term in the adversarial loss.

\citet{arjovsky2019invariant} tackle a similar problem, but take a very different approach,
which they term ``Invariant Risk Minimisation'', contrasted to the usual approach of \acf{ERM}.
As with \citet{kim2019learning},
the goal is to ignore spurious correlations that only appear in the very imperfect training set,
but not in the test set;
and they also perform experiments on Coloured MNIST.
They formalise the problem as one of different environments $e\in E$,
where each environment is a different, biased view of the same underlying data distribution.
The idea is to train a predictor that is simultaneously optimal for all the environments,
with the expectation that this generalises to the test set.
This idea is formulated as a (intractable) bi-level, constrained optimisation problem,
which they approximate with a tractable regularised optimisation.

As the authors point out,
the goal of machine learning should be to identify natural, deep, robust structures in reality,
instead of relying on superficial correlations.
(This dichotomy is sometimes framed as \emph{correlation} vs \emph{causality},
where ``causality'' takes on a very broad meaning that encompasses any kind of fundamental structure in reality.)
Under this view, methods for invariance learning can be framed as trying to learn a more fundamental structure
than the dataset at first seems to show.
By aiming to be invariant to specific environments/subgroups/domains (denoted by the variable \(s\)),
they can be seen as making the claim that \(s\) does not represent a natural structure,
but is merely an artefact of the data generation process.
Furthermore, the ground-truth-centric view of dataset bias can be understood as
aiming to identify the fundamental structures and aiming to be invariant to everything else.
This also fits with \citet{kim2019learning}'s work on Coloured MNIST:
colour is not regarded as the fundamental structure in the data.

@@

A conceptually very influential work was \citet{friedler2016possibility}.
While this work does not explicitly present a ground-truth-centric approach to bias,
their concepts of the ``construct space'' and the ``observed space'' are close
to the idea of the (biased) training distribution and the true underlying distribution.

@@

\citet{kallus2018residual} is another more conceptual work.
They present an intuitive model for how sampling bias can enter a training set
and how it can be very hard to correct for such a bias.
The paper refers to the process as \emph{systematic censoring},
but this is not meant to necessarily imply that this censoring was a conscious decision by some authority;
it can also be an unintended side effect of an enacted policy.
\emph{Systematic censoring} can arise any time a screening process prevents observing the outcome for the screened-out samples.
For example, if, historically, a certain demographic group was screened out from receiving loans,
then it was not possible to observe default rates for this group;
and so any historical data on defaulting is useless for that demographic group.
Even if a bank wants to change their policy,
it is hard to correct for the bias in the training set because there is no good data to learn from.
This is a prime example of a ground-truth-centric bias problem,
as defined in the beginning of the chapter.

Finally, as mentioned in the beginning,
\citet{blum2020recovering} provide a formalisation of a setting with label bias and sampling bias
and an unbiased ground-truth test set.
They explicitly state not being affected by a fairness-accuracy trade-off,
because the true distribution in the considered problem satisfies \acf{DP}.
Their formalised dataset generation process has multiple steps:
First, a small amount of random noise is applied to the (true) labels;
this models general inaccuracies in the labelling process and does not introduce bias yet.
Second, sampling bias is added, by dropping samples based on what the sample's \(s\) and \(y\) values are.
Concretely, only samples with \(s=0\) are dropped.
Finally, labelling bias is added, by flipping labels for one of the demographic groups (\eg, \(s=0\)),
in one specific direction (from \(y=1\) to \(y=0\)).
The authors show that in this specific situation,
an unbiased classifier (as determined by evaluation on the unbiased test set) can be recovered
by enforcing its predictions to be compatible with \acf{EOpp} on the training set.
Notably, the fairness constraint enforced during training (\ac{EOpp})
is not the fairness constraint satisfied by the predictions on the unbiased test set (\ac{DP}).
One reason, why enforcing \ac{EOpp} works so well here is
that the label bias only affects the \emph{underrepresented} demographic group,
and that the label bias is the direction (\(y=1\) to \(y=0\)) that \ac{EOpp} is most sensitive to.
Thus, even in the ground-truth-centric view of bias,
fairness constraints can be very useful.
